---
title: "Inflation_eda"
author: "Dan Gunn"
date: "2024-02-15"
output: html_document
---
##Packages
```{r echo=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(slider) #for moving averages
```

###Loading data

##PCEPI data
Loading data and renaming columns.
PCEPI begins at 100 in base year 2017

Suggested citation: U.S. Bureau of Economic Analysis, "Table 2.8.7. Percent Change From Preceding Period in Prices for Personal Consumption Expenditures by Major Type of Product, Monthly" (accessed Saturday, February 17, 2024).

```{r echo=FALSE}

pcepi_raw_df <- read_csv("raw_data/PCEPI_Numbers 2019-2023/pcepi-Table 1.csv")

#merging month and year together
pcepi_df<- pcepi_raw_df %>%
  mutate(month_num = match(tolower(Month), tolower(month.abb)), # Convert month names to numbers
    date = make_date(Year, month_num, 1) # Create a date with the first day of each month
  )%>%
  select(-Year, -Month, -month_num)

```
Adding the monthly percent change
```{r}
pcepi_df <- pcepi_df %>% 
  mutate(all_items_monthly = (all_items - lag(all_items)) / lag(all_items),
         core_items_monthly = (core_items - lag(core_items)) / lag(core_items)) %>% 
  mutate(all_items_sma = slide_dbl(all_items_monthly, mean, .before = 3, .complete = TRUE),
         core_items_sma = slide_dbl(core_items_monthly, mean, .before = 3, .complete = TRUE))

```

##PPI
Reading in PPI data and changing column names. 
+ Adding a monthly percent change column.
PPI begins at 100 in base year 2012

```{r echo=FALSE}
ppi_data_raw <- read_csv("raw_data/PPI/BLS Data Series-1-Table 1.csv")

long_ppi_data <-ppi_data_raw %>%
  gather(key = "Month", value = "Value", -Year) %>%
  mutate(
    Month = match(Month, month.abb), # Convert month abbreviations to numbers
    Date = as.Date(paste(Year, Month, "01", sep = "-")) # Merge Year and Month into a Date
  ) %>%
  select(Date, Value) %>%
  arrange(Date)# Select only the columns you want

```


```{r}
summary(long_ppi_data)
head(long_ppi_data)
```
Adding the monthly percent change
```{r}
long_ppi_data <- long_ppi_data %>% 
  mutate(ppi_monthly = (Value - lag(Value)) / lag(Value))%>%
  mutate(ppi_sma = slide_dbl(ppi_monthly, mean, .before = 3, .complete = TRUE)) %>%
  rename(date = Date)

```
## CPI
Adding in the Consumer Price Index
All items in U.S. city average, all urban consumers, seasonally adjusted
Series ID	:	CUSR0000SA0
Seasonality	:	Seasonally Adjusted
Base year of 1984

```{r echo = FALSE}
cpi_raw_df <- read_csv("raw_data/DataFinder-20240217201453/cpi-Table 1.csv") %>%
  select_if(~ any(!is.na(.)))

cpi_df <- cpi_raw_df %>%
  mutate(date = as.Date(paste(Year, Month, "1", sep = "-"), format = "%Y-%m-%d")) %>%
  select(-Year, -Month, -Label)%>%
  rename( cpi= Cpi)

cpi_df <- cpi_df %>%
  mutate(cpi_monthly = (cpi - lag(cpi)) / lag(cpi))%>%
  mutate(cpi_sma = slide_dbl(cpi_monthly, mean, .before = 3, .complete = TRUE))
```
Creating a normalized version of the indexes in order to plot them together.

```{r}
all_sets_df <- merge(long_ppi_data, pcepi_df, by = "date")

all_sets_df <- merge(all_sets_df, cpi_df, by = "date")

# base_ppi <- all_sets_df$Value[all_sets_df$date == as.Date('2019-01-01')]
# base_all <- all_sets_df$all_items[all_sets_df$date == as.Date('2019-01-01')]
# base_core <- all_sets_df$core_items[all_sets_df$date == as.Date('2019-01-01')]
# base_cpi <- all_sets_df$cpi[all_sets_df$date == as.Date('2019-01-01')]

specific_date <- as.Date('2019-01-01')

base_values <- all_sets_df %>%
  filter(date == specific_date) %>%
  summarise(across(c(Value, all_items, core_items, cpi), ~ ifelse(is.na(.), NA_real_, .))) %>%
  unlist()

all_sets_df <- all_sets_df %>%
  mutate(across(c(Value, all_items, core_items, cpi),
                ~ . - base_values[cur_column()],
                .names = 'zero_based_{.col}')) %>%
  filter(!if_any(c(zero_based_Value, zero_based_all_items, zero_based_core_items, zero_based_cpi), is.na))


all_sets_df <-all_sets_df %>%
  mutate(ppi_norm = (Value - min(Value)) / (max(Value) - min(Value)),
         all_items_norm = (all_items - min(all_items)) / (max(all_items) - min(all_items)),
         core_norm = (core_items - min(core_items)) / (max(core_items) - min(core_items)),
         cpi_norm = (cpi - min(cpi)) / (max(cpi) - min(cpi)))
         
write_csv(all_sets_df, "clean_data/all_sets.csv")
```

Plotting with the normalized values of the index.

```{r}
ggplot(data=all_sets_df)+
  geom_line(aes(x=date, y = all_items_norm))+
  geom_line(aes(x=date, y = core_norm),
            color= "blue") +
  geom_line(aes(x=date, y = ppi_norm),
            color = "red")+
  geom_line(aes(x = date, y = cpi_norm),
            color = "darkgreen")+
  theme_minimal()
```

```{r}
ggplot(data=all_sets_df)+
  geom_line(aes(x=date, y = zero_based_all_items))+
  geom_line(aes(x=date, y = zero_based_core_items),
            color= "blue") +
  geom_line(aes(x=date, y = zero_based_Value),
            color = "red")+
  geom_line(aes(x = date, y = zero_based_cpi),
            color = "darkgreen")+
  theme_minimal()
```


```{r}
ggplot(data=pcepi_df)+
  geom_line(aes(x=date, y = all_items_monthly*100))+
  geom_line(aes(x=date, y = core_items_monthly*100),
            color= "blue")+
   geom_line(data=long_ppi_data, 
             aes(x=date, y = ppi_monthly*100),
             color = "red")+
   geom_line(data = cpi_df,
            aes(x = date, y = cpi_monthly*100),
            color = "darkgreen")+
  theme_minimal()
```

I should do a moving average of these numbers to smooth them out a bit. 

```{r}
ggplot(data=pcepi_df)+
  geom_line(aes(x=date, y = all_items_sma*100))+
  geom_line(aes(x=date, y = core_items_sma*100),
            color= "blue",
            size= 1,
            alpha = 0.5)+
   geom_line(data=long_ppi_data, 
             aes(x=date, y = ppi_sma*100),
             color = "red")+
   geom_line(data = cpi_df,
            aes(x = date, y = cpi_sma*100),
            color = "darkgreen")+
  theme_minimal()
```
Maybe add at Smoothed Moving Average (smma)


##Correlation Testing

```{r}
#PPI to PCEPI Cross-Correlation Function CCF
ccf_values_ppi_pcepi = ccf(long_ppi_data$Value, pcepi_df$core_items, lag.max = 12, main = "Cross-Correlation Function CCF")

ccf_values_ppi_pcepi
```

```{r}
#PPI to CPI Cross-Correlation Function CCF
ccf_values_ppi_cpi = ccf(long_ppi_data$Value, cpi_df$cpi, lag.max = 12, main = "Cross-Correlation Function CCF")

ccf_values_ppi_cpi
```


#Granger Causality testing to see if the PPI really predicts other inflation measures.
```{r}
library(lmtest) #for granger testing
granger_result <- grangertest(pcepi_df$core_items ~ long_ppi_data$Value, order = 1) # 'order' is the number of lags to include

# View the results
print(granger_result)
```
## Akaike Information Criterion (AIC)
Trying to see the optimal order number via Akaike Information Criterion (AIC)

```{r}
combined_df <- merge(long_ppi_data, pcepi_df, by = "date")
# Placeholder for AIC values and corresponding lags
aic_values <- setNames(rep(NA, 6), 1:6)

for (i in 1:6) {
  # Dynamically create 'lagged_Value' within the loop to reflect the current lag being tested
  temp_df <- combined_df %>%
    mutate(lagged_Value = lag(Value, i))

  model <- lm(core_items ~ lagged_Value + lag(core_items, i), data = temp_df)
  aic_values[i] <- AIC(model)
}

# Identify the best lag
best_lag <- which.min(aic_values)
best_aic <- min(aic_values)

print(paste("Best lag for Value and core_items:", best_lag, "with AIC:", best_aic))
```
##Results
The best lag is 1. 
The result implies that the most recent past value of PPI (lag of 1 month) has the most significant predictive power on core inflation items. This suggests that changes in PPI quickly reflect or predict changes in core inflation, with the strongest relationship occurring on a month-to-month basis.

Short-term Dynamics:
Since the optimal lag is 1 month, it indicates short-term dynamics between PPI and core inflation items. The PPI's influence on core inflation items is immediate and most pronounced in the following month, highlighting the responsiveness of core inflation to producer price changes within a short time frame.

```{r}
library(lmtest) #for granger testing
granger_result_cpi <- grangertest(cpi_df$cpi ~ long_ppi_data$Value, order = 12) # 'order' is the number of lags to include

# View the results
print(granger_result_cpi)
```
It also predicts CPI

```{r}
lags <- 1:12
p_values <- c(0.00003602, 0.01117, 0.0002295, 0.002702, 0.005492, 0.009191, 
              0.009291,0.005237, 0.007687,0.007339,0.02733, 0.05778    ) #p-values for each lag
significance_levels <- ifelse(p_values < 0.05, "Significant", "Not Significant")

# Creating a data frame for plotting
data_for_plot <- data.frame(lags, p_values, significance_levels)

# Plotting using ggplot2
library(ggplot2)

ggplot(data_for_plot, aes(x = lags, y = p_values, color = significance_levels)) +
  geom_point() +
  geom_line(aes(group = 1)) +
  scale_y_continuous(trans = 'reverse') + # Optional: Reverse scale for p-values to have lower values at the top
  labs(x = "Lag", y = "P-Value", title = "Granger Causality Test Results by Lag") +
  theme_minimal() +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "blue")) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  annotate("text", x = max(lags) - 1, y = 0.05, label = "Significance Threshold (0.05)", hjust = 1)
```

##Working to predict the CPI or PCEPI from the PPI
3 month lag time. 

```{r}
# Create a lagged PPI column for 6 months prior
combined_df <- combined_df %>%
  arrange(date) %>%
  mutate(lagged_PPI_3m = lag(Value, 3),
         lagged_PPI_2m = lag(Value, 2),
         lagged_PPI_6m = lag(Value, 6))

# Split the data into training and test sets
# Assuming the test set starts from January 2022
training_set <- combined_df %>% filter(date < as.Date("2022-01-01"))
test_set <- combined_df %>% filter(date >= as.Date("2022-01-01"))

# Fit a model using the training set
model <- lm(core_items ~ lagged_PPI_6m + lagged_PPI_3m, data = training_set)

# Generate predictions for the test set
test_set <- test_set %>%
  mutate(predicted_core_items = predict(model, newdata = test_set))

# Combine the actual and predicted values for visualization
visualization_data <- test_set %>%
  select(date, core_items, predicted_core_items)

# Visualize the actual vs. predicted core inflation
ggplot(visualization_data, aes(x = date)) +
  geom_line(aes(y = core_items, color = "Actual")) +
  geom_line(aes(y = predicted_core_items, color = "Predicted")) +
  labs(title = "PPI Predicted vs Actual Core Inflation",
       y = "Core Inflation Index",
       color = "Legend") +
  theme_minimal()
```

